---
title: "Master_Table"
output:
  github_document:
    html_preview: false
    toc: true
    toc_depth: 2
always_allow_html: true
editor_options:
  chunk_output_type: console
knit: (function(input_file, encoding) {
    rmarkdown::render(input_file,output_file= 'Readme.md')
        })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, echo = TRUE)

# Fold every code block into a collapsible <details> element. GitHub renders this
# natively, so the rendered Readme.md reads as a summary while the full code stays
# one click away.
knitr::knit_hooks$set(source = function(x, options) {
  paste0("\n<details>\n<summary>Code</summary>\n\n``` r\n",
         paste(x, collapse = "\n"), "\n```\n\n</details>\n")
})

# All summary output in this document goes through kable, so nothing dumps a
# raw multi-page data.frame into the rendered file. The caption is emitted as a
# bold line *above* the table - kable's own caption ends up below it, where it
# reads as a stray sentence on GitHub.
tab <- function(x, caption = NULL) {
  md <- paste(knitr::kable(x), collapse = "\n")
  if (!is.null(caption)) md <- paste0("**", caption, "**\n\n", md)
  knitr::asis_output(paste0("\n", md, "\n"))
}
```

This document merges the three RNA-seq / microarray data sets of the project into
one **master table**, classifies every gene by its HIF dependency, and exports the
result as Excel. Code blocks are collapsed - click *Code* to expand.

```{r load_packages}
library(dplyr)
library(DESeq2)
library(tidyverse)
library(knitr)
library(kableExtra)
library(ggplot2)
library(ggrepel)
library(openxlsx)
library(VennDiagram)
library(patchwork)

# Suppress the .log files the VennDiagram package writes on every call.
invisible(futile.logger::flog.threshold(futile.logger::ERROR, name = "VennDiagramLogger"))

colors <- c("lavenderblush3", "lavenderblush4", "#90caf9", "#1976d2",
            "#82e0aa", "#239b56", "#f8c471", "#b9770e")

```


# Input data

Three sources go into the master table: the Kelly line with its three HIF knockouts,
the two MYCN-non-amplified reference lines, and the high-risk microarray gene list.

```{r load_data}

dds_kelly <- readRDS(
  file = "/Users/simonkelterborn/Library/CloudStorage/GoogleDrive-simon.kelterborn@gmail.com/Meine Ablage/! OneDrive Charité/dds_filtered_with_metadata.rds"
)

load(
  file = "/Users/simonkelterborn/Library/CloudStorage/GoogleDrive-simon.kelterborn@gmail.com/Meine Ablage/! OneDrive Charité/SKNAS_SHSY5Y.dds"
)
dds_SKNAS <- dds
rm(dds)

HR_table <- readxl::read_xlsx("/Users/simonkelterborn/Library/CloudStorage/GoogleDrive-simon.kelterborn@gmail.com/Meine Ablage/! OneDrive Charité/HR_genes_Microarray.xlsx")

tab(data.frame(
  Data_set = c("Kelly (WT + HIF1a/HIF1b/HIF2a KO)",
               "SK-N-AS / SH-SY5Y",
               "High-risk microarray"),
  File     = c("dds_filtered_with_metadata.rds", "SKNAS_SHSY5Y.dds",
               "HR_genes_Microarray.xlsx"),
  Genes    = c(nrow(dds_kelly), nrow(dds_SKNAS), nrow(HR_table)),
  Samples  = c(ncol(dds_kelly), ncol(dds_SKNAS), NA)
), "Input data sets")

```


# Kelly: fresh contrasts from the new dds

DESeq2 is run once with the full interaction design `~genotype + condition +
genotype:condition`. All contrasts of interest are then derived from that single
fit as weighted coefficient vectors, so every comparison shares one dispersion
estimate. The coefficient names are resolved by pattern instead of being
hardcoded, because DESeq2 rewrites factor level separators (`HIF1a-KO` becomes
`HIF1a.KO`).

```{r kelly_contrasts}

colData(dds_kelly)$genotype <- relevel(factor(colData(dds_kelly)$genotype), "WT")
colData(dds_kelly)$condition <- relevel(factor(colData(dds_kelly)$condition), "Nx")

design(dds_kelly) <- ~genotype + condition + genotype:condition
dds_kelly <- DESeq(dds_kelly)

# entrezid is a 1-to-many list-column (some genes map to multiple Entrez IDs) -
# flatten to a single string so results/data.frame conversions don't explode rows.
entrez_flat <- vapply(mcols(dds_kelly)$entrezid, function(x) {
  if (length(x) == 0) NA_character_ else paste(x, collapse = ";")
}, character(1))

rn <- resultsNames(dds_kelly)

# Resolve the actual (possibly sanitized, e.g. "HIF1a-KO" -> "HIF1a.KO") coefficient
# names by pattern rather than hardcoding them, since DESeq2 rewrites factor level
# separators and level order can change if the input dds changes.
find_coef <- function(pattern) {
  hits <- grep(pattern, rn, value = TRUE)
  if (length(hits) != 1) {
    stop("Expected exactly 1 match for pattern '", pattern, "', found: ",
         paste(hits, collapse = ", "))
  }
  hits
}

coef_geno <- list(
  Hif1a = find_coef("^genotype.*HIF1a.*_vs_WT$"),
  Hif1b = find_coef("^genotype.*HIF1b.*_vs_WT$"),
  Hif2a = find_coef("^genotype.*HIF2a.*_vs_WT$")
)
coef_cond <- find_coef("^condition_Hx_vs_Nx$")
coef_int <- list(
  Hif1a = find_coef("^genotype.*HIF1a.*\\.condition"),
  Hif1b = find_coef("^genotype.*HIF1b.*\\.condition"),
  Hif2a = find_coef("^genotype.*HIF2a.*\\.condition")
)

# Named weight vectors, resolved to the real coefficient names above.
# Mirrors the full contrast set from 2B_DGE/2B_DGE.Rmd, computed fresh here.
contrast_defs <- list(
  Kelly.Hx.vs.Nx = c(setNames(1, coef_cond)),
  Hif1a.Hx.vs.Nx = c(setNames(1, coef_cond), setNames(1, coef_int$Hif1a)),
  Hif1b.Hx.vs.Nx = c(setNames(1, coef_cond), setNames(1, coef_int$Hif1b)),
  Hif2a.Hx.vs.Nx = c(setNames(1, coef_cond), setNames(1, coef_int$Hif2a)),

  Nx.Hif1a.vs.Kelly = c(setNames(1, coef_geno$Hif1a)),
  Nx.Hif1b.vs.Kelly = c(setNames(1, coef_geno$Hif1b)),
  Nx.Hif2a.vs.Kelly = c(setNames(1, coef_geno$Hif2a)),

  Hx.Hif1a.vs.Kelly = c(setNames(1, coef_geno$Hif1a), setNames(1, coef_int$Hif1a)),
  Hx.Hif1b.vs.Kelly = c(setNames(1, coef_geno$Hif1b), setNames(1, coef_int$Hif1b)),
  Hx.Hif2a.vs.Kelly = c(setNames(1, coef_geno$Hif2a), setNames(1, coef_int$Hif2a)),

  Hx.Hif2a.vs.Hif1a = c(setNames(-1, coef_geno$Hif1a), setNames(-1, coef_int$Hif1a),
                        setNames(1, coef_geno$Hif2a), setNames(1, coef_int$Hif2a)),
  Hx.Hif1b.vs.Hif1a = c(setNames(-1, coef_geno$Hif1a), setNames(-1, coef_int$Hif1a),
                        setNames(1, coef_geno$Hif1b), setNames(1, coef_int$Hif1b)),
  Hx.Hif1b.vs.Hif2a = c(setNames(-1, coef_geno$Hif2a), setNames(-1, coef_int$Hif2a),
                        setNames(1, coef_geno$Hif1b), setNames(1, coef_int$Hif1b)),

  Hif1aHxNx.vs.KellyHxNx = c(setNames(1, coef_int$Hif1a)),
  Hif2aHxNx.vs.KellyHxNx = c(setNames(1, coef_int$Hif2a)),
  Hif1bHxNx.vs.KellyHxNx = c(setNames(1, coef_int$Hif1b)),
  Hif2aHxNx.vs.Hif1aHxNx = c(setNames(-1, coef_int$Hif1a), setNames(1, coef_int$Hif2a)),

  Hx.Hif1b.vs.Hif12a = c(setNames(-0.5, coef_geno$Hif1a), setNames(-0.5, coef_geno$Hif2a),
                         setNames(1, coef_geno$Hif1b),
                         setNames(-0.5, coef_int$Hif1a), setNames(-0.5, coef_int$Hif2a),
                         setNames(1, coef_int$Hif1b)),

  Hx.Kelly.vs.allHIFs = c(setNames(1/3, coef_geno$Hif1a), setNames(1/3, coef_geno$Hif2a),
                          setNames(1/3, coef_geno$Hif1b),
                          setNames(1/3, coef_int$Hif1a), setNames(1/3, coef_int$Hif2a),
                          setNames(1/3, coef_int$Hif1b)),

  # NOTE: kept "NOT confirmed" caveat from the original 2B_DGE.Rmd analysis
  Hx.vs.Nx = c(setNames(1, coef_cond),
               setNames(1/4, coef_int$Hif1a), setNames(1/4, coef_int$Hif2a),
               setNames(1/4, coef_int$Hif1b))
)

build_contrast_vec <- function(weights) {
  v <- setNames(rep(0, length(rn)), rn)
  v[names(weights)] <- unname(weights)
  unname(v)
}

results_list <- lapply(contrast_defs, function(w) {
  r1 <- results(dds_kelly, contrast = build_contrast_vec(w))
  r1$symbol  <- mcols(dds_kelly)$symbol
  r1$ENSEMBL <- mcols(dds_kelly)$gene_id
  r1$ENTREZ  <- entrez_flat
  r1
})

tab(data.frame(
  Contrast = names(results_list),
  DEGs     = sapply(results_list, function(x) sum(x$padj < 0.05, na.rm = TRUE)),
  row.names = NULL
), "Significant genes per contrast (padj < 0.05)")

```


# SK-N-AS / SH-SY5Y: independent per-line Hx vs Nx

These two lines are analysed separately from Kelly - each gets its own DESeq2 fit
on the subset of its own samples with the simple design `~Condition`, so their
hypoxia response is estimated independently of the Kelly model.

```{r sknas_shsy5y_contrasts}

get_line_hxvsnx <- function(dds_all, celltype_value, label) {
  dds_sub <- dds_all[, colData(dds_all)$Celltype == celltype_value]
  colData(dds_sub)$Condition <- relevel(factor(colData(dds_sub)$Condition), "Normoxia")
  design(dds_sub) <- ~Condition
  dds_sub <- DESeq(dds_sub)
  res <- results(dds_sub, name = "Condition_Hypoxia_vs_Normoxia")

  out <- data.frame(
    Ensembl = rownames(res),
    baseMean = res$baseMean,
    log2FC   = res$log2FoldChange,
    padj     = res$padj
  )
  colnames(out)[2:4] <- paste0(c("baseMean_", "log2FC_", "padj_"), label)
  out
}

sknas_df  <- get_line_hxvsnx(dds_SKNAS, "SK-N-AS", "SKNAS")
shsy5y_df <- get_line_hxvsnx(dds_SKNAS, "SH-SY5Y", "SHSY5Y")

```


# Combine into master table

All three sources are joined on Ensembl gene ID, with Kelly as the reference gene
space. Every contrast contributes a `log2FoldChange` and a `padj` column.

```{r build_master_table}

res_table <- lapply(results_list, function(x) data.frame(x)[, c("log2FoldChange", "padj")])
res_table <- do.call('cbind', res_table)

master_table <- data.frame(
  Ensembl        = mcols(dds_kelly)$gene_id,
  ENTREZ         = entrez_flat,
  symbol         = mcols(dds_kelly)$symbol,
  baseMean_Kelly = results_list[["Kelly.Hx.vs.Nx"]]$baseMean
)

master_table <- cbind(master_table, res_table)

master_table <- master_table %>%
  left_join(sknas_df, by = "Ensembl") %>%
  left_join(shsy5y_df, by = "Ensembl") %>%
  left_join(HR_table %>% dplyr::rename(p_value_HR = p_value), by = "Ensembl")

```

## Overlap of the Ensembl IDs

How well the three sources cover the Kelly gene space:

```{r ensembl_overlap}

ens_kelly <- master_table$Ensembl

tab(data.frame(
  Data_set          = c("Kelly (reference)", "SK-N-AS / SH-SY5Y", "High-risk microarray"),
  Genes             = c(length(ens_kelly), nrow(sknas_df), nrow(HR_table)),
  Shared_with_Kelly = c(NA, sum(sknas_df$Ensembl %in% ens_kelly),
                        sum(HR_table$Ensembl %in% ens_kelly)),
  Not_in_Kelly      = c(NA, sum(!sknas_df$Ensembl %in% ens_kelly),
                        sum(!HR_table$Ensembl %in% ens_kelly))
), "Ensembl ID overlap with the Kelly gene space")

```

Being present is not the same as being testable: DESeq2's independent filtering
drops low-count genes, which leaves their `padj` as `NA`. That distinction matters
for the baseline classification below, so it is counted separately here.

```{r baseline_coverage}

coverage <- function(label) {
  bm <- master_table[[paste0("baseMean_", label)]]
  pa <- master_table[[paste0("padj_", label)]]
  data.frame(Line                     = label,
             Not_in_data_set          = sum(is.na(bm)),
             Present_but_not_testable = sum(!is.na(bm) & is.na(pa)),
             Testable                 = sum(!is.na(pa)))
}

tab(rbind(coverage("SKNAS"), coverage("SHSY5Y")),
    "Gene coverage of the two baseline lines")

```

```{r master_table_preview}

# p-values are rendered as text: kable applies one numeric format across a whole
# column, which turns anything below ~1e-4 into a misleading "0.0000".
master_table %>%
  dplyr::select(Ensembl, symbol, baseMean_Kelly,
                Kelly.Hx.vs.Nx.log2FoldChange, Kelly.Hx.vs.Nx.padj,
                log2FC_SKNAS, log2FC_SHSY5Y) %>%
  head(5) %>%
  dplyr::mutate(Kelly.Hx.vs.Nx.padj = sprintf("%.1e", Kelly.Hx.vs.Nx.padj),
                dplyr::across(where(is.numeric), ~signif(.x, 3))) %>%
  tab(sprintf("master_table preview - first 5 of %s genes, 7 of %d columns",
              format(nrow(master_table), big.mark = ","), ncol(master_table)))

```


# Target classification (Kelly only)

A gene counts as a **HIF target** if it is induced by hypoxia in WT Kelly
(`padj < 0.05` and `log2FC > 1`) *and* that induction is lost in the respective
knockout, i.e. the interaction term is significantly negative (`padj < 0.05` and
`log2FC < -1`). Genes meeting both criteria for HIF1A and HIF2A are labelled
`Hif1a_Hif2a`. Only Kelly data is used here.

```{r target_column}

padj_cut <- 0.05
lfc_cut  <- 1

sig_up   <- function(lfc, padj) !is.na(padj) & padj < padj_cut & !is.na(lfc) & lfc >  lfc_cut
sig_down <- function(lfc, padj) !is.na(padj) & padj < padj_cut & !is.na(lfc) & lfc < -lfc_cut

hx_up <- sig_up(master_table$Kelly.Hx.vs.Nx.log2FoldChange,
                master_table$Kelly.Hx.vs.Nx.padj)

is_hif1a <- hx_up & sig_down(master_table$Hif1aHxNx.vs.KellyHxNx.log2FoldChange,
                             master_table$Hif1aHxNx.vs.KellyHxNx.padj)
is_hif2a <- hx_up & sig_down(master_table$Hif2aHxNx.vs.KellyHxNx.log2FoldChange,
                             master_table$Hif2aHxNx.vs.KellyHxNx.padj)

master_table$Target <- dplyr::case_when(
  is_hif1a & is_hif2a ~ "Hif1a_Hif2a",
  is_hif1a            ~ "Hif1a",
  is_hif2a            ~ "Hif2a",
  TRUE                ~ NA_character_
)

master_table <- master_table %>% dplyr::relocate(Target, .after = symbol)

# Explicit labels + level order, so NA does not surface as the string "NA." and
# the categories read in a sensible order rather than alphabetically.
target_f <- factor(ifelse(is.na(master_table$Target), "no target", master_table$Target),
                   levels = c("Hif1a", "Hif2a", "Hif1a_Hif2a", "no target"))

tab(as.data.frame(table(Target = target_f)),
    sprintf("Target classification (%s genes Hx-induced in WT Kelly)",
            format(sum(hx_up), big.mark = ",")))

```


# Hypoxia baseline (SK-N-AS / SH-SY5Y)

SK-N-AS and SH-SY5Y are both **MYCN-non-amplified** neuroblastoma lines and serve
as the baseline hypoxia response; Kelly is MYCN-amplified. The same up-criterion
as for `Target` is applied, so both columns are directly comparable.

| Value | Meaning |
|---|---|
| `SKNAS_SHSY5Y` | Hx-induced in both baseline lines |
| `SKNAS` / `SHSY5Y` | Hx-induced in that line only |
| `none` | testable in at least one line, but not Hx-induced - the Kelly-specific candidates |
| `NA` | testable in neither line - unknown, **not** "no response" |

Keeping `NA` separate from `none` matters: without it, genes that were never
measured in the reference lines would masquerade as Kelly-specific.

```{r baseline_column}

up_sknas  <- sig_up(master_table$log2FC_SKNAS,  master_table$padj_SKNAS)
up_shsy5y <- sig_up(master_table$log2FC_SHSY5Y, master_table$padj_SHSY5Y)

testable <- !is.na(master_table$padj_SKNAS) | !is.na(master_table$padj_SHSY5Y)

master_table$Hx_Baseline <- dplyr::case_when(
  up_sknas & up_shsy5y ~ "SKNAS_SHSY5Y",
  up_sknas             ~ "SKNAS",
  up_shsy5y            ~ "SHSY5Y",
  testable             ~ "none",
  TRUE                 ~ NA_character_
)

master_table <- master_table %>% dplyr::relocate(Hx_Baseline, .after = Target)

baseline_f <- factor(ifelse(is.na(master_table$Hx_Baseline), "no data",
                            master_table$Hx_Baseline),
                     levels = c("none", "SKNAS", "SHSY5Y", "SKNAS_SHSY5Y", "no data"))

tab(as.data.frame(table(Hx_Baseline = baseline_f)), "Baseline hypoxia response")

```

## Target vs. baseline

Rows are the HIF targets from Kelly, columns their behaviour in the
MYCN-non-amplified lines. The `none` column holds the Kelly-specific targets, the
`NA` column the ones with no baseline data.

```{r target_baseline_crosstable}

cross <- table(Target = target_f, Baseline = baseline_f)

tab(as.data.frame.matrix(cross),
    sprintf("Target x Hx_Baseline - %s HIF targets are Kelly-specific",
            format(sum(!is.na(master_table$Target) &
                       master_table$Hx_Baseline == "none", na.rm = TRUE),
                   big.mark = ",")))

```

## Venn: Target vs. Hx_Baseline

Left: both reference lines pooled into one set. Right: the same comparison with
SK-N-AS and SH-SY5Y resolved separately.

```{r target_baseline_venn, fig.width=10, fig.height=5, dpi=150}

# Restrict to genes that are testable in at least one baseline line. Otherwise
# genes with no baseline data would sit outside the SKNAS/SHSY5Y circles and
# would be read as "Kelly-specific" when they are simply unmeasured.
venn_universe <- master_table %>% dplyr::filter(!is.na(Hx_Baseline))

ens_where <- function(keep) venn_universe$Ensembl[keep]

venn_sets <- list(
  Hif1a  = ens_where(venn_universe$Target %in% c("Hif1a", "Hif1a_Hif2a")),
  Hif2a  = ens_where(venn_universe$Target %in% c("Hif2a", "Hif1a_Hif2a")),
  SKNAS  = ens_where(venn_universe$Hx_Baseline %in% c("SKNAS", "SKNAS_SHSY5Y")),
  SHSY5Y = ens_where(venn_universe$Hx_Baseline %in% c("SHSY5Y", "SKNAS_SHSY5Y"))
)

venn_plot <- function(input_list, main, fill) {
  venn.diagram(
    x = input_list,
    category.names = paste0(names(input_list), "\n(", lengths(input_list), ")"),
    force.unique = TRUE, na = "remove",
    filename = NULL,
    main = main, main.fontface = "bold",
    lwd = 2,
    lty = "blank",
    fill = fill,
    cat.fontface = "bold",
    cex = 0.8, cat.cex = 0.8
  )
}

# Simplified view: both baseline lines pooled into one "Hx in baseline" set.
venn_simple <- list(
  Hif1a = venn_sets$Hif1a,
  Hif2a = venn_sets$Hif2a,
  Baseline_Hx = union(venn_sets$SKNAS, venn_sets$SHSY5Y)
)

plt_simple <- venn_plot(venn_simple, "HIF targets vs. baseline Hx response",
                        colors[c(4, 6, 1)])
plt_full   <- venn_plot(venn_sets, "HIF targets vs. SK-N-AS / SH-SY5Y",
                        colors[c(4, 6, 7, 3)])

wrap_elements(plt_simple) + wrap_elements(plt_full)

```

```{r venn_universe_note, echo=FALSE, results='asis'}

cat(sprintf("*Venn universe: %s genes testable in at least one baseline line, %s genes excluded for lack of baseline data.*\n",
            format(nrow(venn_universe), big.mark = ","),
            format(sum(is.na(master_table$Hx_Baseline)), big.mark = ",")))

```


# Literature coverage

Three columns describing how well a gene is already studied. The counts come from
NCBI's curated `gene2pubmed` mapping (Entrez GeneID -> PubMed ID) rather than from
per-gene PubMed text searches - one download instead of 64,537 API calls, and no
dependence on symbol matching.

- `pubmed_total` - curated publications for the gene
- `pubmed_hypoxia` - of those, publications in the hypoxia / HIF literature
- `pubmed_source` - how the gene was matched to an Entrez ID

The second column is the useful one here. `pubmed_total` mostly measures how long
a gene has been known; `pubmed_hypoxia` answers whether *this* biology is already
described. A strongly induced, Kelly-specific target with many publications but
none in hypoxia is the interesting case.

**Why three ID sources.** The `ENTREZ` column of our annotation leaves 35,994
genes unmapped, and a quarter of the HIF targets sit in that gap - too many to
ignore for a column meant to help prioritise. Two NCBI tables close most of it
without giving up precision: `gene2ensembl` matches on the Ensembl gene ID, and
`Homo_sapiens.gene_info` on primary symbols and synonyms. Any key pointing at
more than one gene is discarded, so unlike a `"SYMBOL"[Gene Name]` text search
this cannot mismatch `CS`, `AR`, `MARCH1` or `SEPT9`.

`pubmed_source` records which route was used, so the value stays usable as a soft
ranking aid while its provenance is always visible. Genes marked `unannotated`
have neither an Entrez ID nor a symbol - there is no identifier to look literature
up by, so they stay `NA` rather than being called unstudied. Only genes that were
mapped and have no linked publication get a real `0`.

```{r pubmed_coverage}

cache_dir <- here::here("6_XM_data_merge", "cache")
dir.create(cache_dir, showWarnings = FALSE, recursive = TRUE)

ncbi <- c(gene2pubmed  = "https://ftp.ncbi.nlm.nih.gov/gene/DATA/gene2pubmed.gz",
          gene2ensembl = "https://ftp.ncbi.nlm.nih.gov/gene/DATA/gene2ensembl.gz",
          gene_info    = paste0("https://ftp.ncbi.nlm.nih.gov/gene/DATA/GENE_INFO/",
                                "Mammalia/Homo_sapiens.gene_info.gz"))
files <- file.path(cache_dir, c("gene2pubmed.gz", "gene2ensembl.gz",
                                "Homo_sapiens.gene_info.gz"))
names(files) <- names(ncbi)
for (n in names(files)) {
  if (!file.exists(files[n])) download.file(ncbi[n], files[n], mode = "wb", quiet = TRUE)
}
pmid_file <- file.path(cache_dir, "hypoxia_pmids.rds")

# Human subset only; the full files are tens of millions of rows, so filter while
# decompressing rather than reading everything into memory.
g2p <- data.table::fread(
  cmd = sprintf("gzcat %s | awk -F'\t' '$1==9606'", shQuote(files["gene2pubmed"])),
  header = FALSE, col.names = c("tax_id", "GeneID", "PubMed_ID"))

g2e <- data.table::fread(
  cmd = sprintf("gzcat %s | awk -F'\t' '$1==9606'", shQuote(files["gene2ensembl"])),
  header = FALSE, select = 1:3, col.names = c("tax_id", "GeneID", "Ensembl"))
g2e <- unique(g2e[, .(Ensembl, GeneID)])

# Symbol -> GeneID via NCBI's own table (primary symbols first, then synonyms).
# Any key pointing at more than one gene is dropped, so this cannot mismatch the
# way a "SYMBOL"[Gene Name] text search would on CS, AR, MARCH1 or SEPT9.
gi <- data.table::fread(cmd = sprintf("gzcat %s", shQuote(files["gene_info"])),
                        quote = "", select = c(2, 3, 5),
                        col.names = c("GeneID", "Symbol", "Synonyms"))
# NB: the column is called "sym", not "key" - data.table() treats a "key" argument
# as the key to set rather than as a column.
prim <- unique(gi[Symbol != "-", .(sym = Symbol, GeneID)])
prim <- prim[!sym %in% prim[, .N, by = sym][N > 1, sym]]
syn <- gi[Synonyms != "-"][, .(sym = unlist(strsplit(Synonyms, "|", fixed = TRUE))),
                           by = GeneID][, .(sym, GeneID)]
syn <- unique(syn[!sym %in% prim$sym])
syn <- syn[!sym %in% syn[, .N, by = sym][N > 1, sym]]
sym_map <- rbind(prim, syn)

hypoxia_query <- paste0(
  'hypoxia[Title/Abstract] OR "hypoxia-inducible factor"[Title/Abstract] ',
  'OR HIF-1[Title/Abstract] OR HIF-2[Title/Abstract] OR hypoxia[MeSH Terms]')

if (file.exists(pmid_file)) {
  hyp_ids <- readRDS(pmid_file)
} else {
  # esearch and efetch both cap at 10,000 records per request, so the result set
  # is split by publication year, and years above the cap by month.
  qenc <- URLencode(paste0("(", hypoxia_query, ")"), reserved = TRUE)
  esearch <- function(mind, maxd, retmax) {
    u <- paste0("https://eutils.ncbi.nlm.nih.gov/entrez/eutils/esearch.fcgi",
                "?db=pubmed&retmode=json&retmax=", retmax, "&term=", qenc,
                "&datetype=pdat&mindate=", mind, "&maxdate=", maxd)
    for (att in 1:4) {
      j <- tryCatch(jsonlite::fromJSON(u), error = function(e) NULL)
      if (!is.null(j) && !is.null(j$esearchresult$count)) return(j$esearchresult)
      Sys.sleep(1.5)
    }
    stop("esearch failed for ", mind, "-", maxd)
  }
  window_ids <- function(mind, maxd) {
    n <- as.integer(esearch(mind, maxd, 0)$count)
    if (n == 0) return(character(0))
    if (n > 9500) return(NULL)
    Sys.sleep(0.36)
    as.character(esearch(mind, maxd, 10000)$idlist)
  }
  hyp_ids <- character(0)
  for (y in 1945:as.integer(format(Sys.Date(), "%Y"))) {
    Sys.sleep(0.36)
    got <- window_ids(y, y)
    if (is.null(got)) {
      for (m in 1:12) {
        Sys.sleep(0.36)
        md <- sprintf("%d/%02d", y, m)
        got_m <- window_ids(md, md)
        if (is.null(got_m)) stop("month window still above the cap: ", md)
        hyp_ids <- c(hyp_ids, got_m)
      }
    } else {
      hyp_ids <- c(hyp_ids, got)
    }
  }
  hyp_ids <- unique(as.integer(hyp_ids))
  saveRDS(hyp_ids, pmid_file)
}

# openxlsx writes NA_character_ as the literal string "NA", so a table that has
# been through the Excel round trip carries text where a missing value belongs.
# Normalise before anything keys off is.na().
entrez <- master_table$ENTREZ
entrez[entrez %in% c("NA", "")] <- NA_character_

# Resolve an Entrez ID per gene from three sources in decreasing directness. The
# annotation alone leaves 36k genes unmapped, a quarter of them inside the HIF
# target set, so the two NCBI fallbacks are worth the extra join.
DT <- data.table::data.table
n_genes <- nrow(master_table)

from_annot <- DT(row_id = seq_len(n_genes), ENTREZ = entrez)[!is.na(ENTREZ)]
from_annot <- from_annot[, .(GeneID = as.integer(unlist(strsplit(ENTREZ, ";")))),
                         by = row_id][!is.na(GeneID)]

open_rows <- setdiff(seq_len(n_genes), from_annot$row_id)
from_ens <- merge(DT(row_id = open_rows, Ensembl = master_table$Ensembl[open_rows]),
                  g2e, by = "Ensembl")[, .(row_id, GeneID)]

open_rows <- setdiff(open_rows, from_ens$row_id)
from_sym <- merge(DT(row_id = open_rows, sym = master_table$symbol[open_rows])[!is.na(sym)],
                  sym_map, by = "sym")[, .(row_id, GeneID)]

map <- rbind(from_annot, from_ens, from_sym)

# Provenance, so the number stays usable as a soft ranking aid while it is always
# visible how it was obtained. "unannotated" genes carry neither an Entrez ID nor
# a symbol - no identifier exists to look literature up by, so they stay NA.
pubmed_source <- rep("unannotated", n_genes)
pubmed_source[from_sym$row_id]   <- "symbol_map"
pubmed_source[from_ens$row_id]   <- "ensembl_map"
pubmed_source[from_annot$row_id] <- "annotation"
master_table$pubmed_source <- factor(
  pubmed_source, levels = c("annotation", "ensembl_map", "symbol_map", "unannotated"))

# Take the UNION of PMIDs across a gene's Entrez IDs, so multi-mapping genes are
# not counted twice.
hit <- merge(map, g2p[, .(GeneID, PubMed_ID)], by = "GeneID", allow.cartesian = TRUE)
hit <- unique(hit[, .(row_id, PubMed_ID)])
hit[, is_hyp := PubMed_ID %in% hyp_ids]
agg <- hit[, .(total = .N, hypoxia = sum(is_hyp)), by = row_id]

mapped <- pubmed_source != "unannotated"
master_table$pubmed_total   <- ifelse(mapped, 0L, NA_integer_)
master_table$pubmed_hypoxia <- ifelse(mapped, 0L, NA_integer_)
master_table$pubmed_total[agg$row_id]   <- agg$total
master_table$pubmed_hypoxia[agg$row_id] <- agg$hypoxia

master_table <- master_table %>%
  dplyr::relocate(pubmed_total, pubmed_hypoxia, pubmed_source, .after = Hx_Baseline)

tab(as.data.frame(table(Source = master_table$pubmed_source)),
    sprintf("How the Entrez ID was resolved (%s hypoxia PMIDs, %s gene-paper links)",
            format(length(hyp_ids), big.mark = ","), format(nrow(g2p), big.mark = ",")))

tab(data.frame(
  Group = c("mapped, no publication", "at least 1 publication",
            "at least 1 hypoxia publication", "unannotated (NA - no identifier)"),
  Genes = c(sum(master_table$pubmed_total == 0, na.rm = TRUE),
            sum(master_table$pubmed_total > 0, na.rm = TRUE),
            sum(master_table$pubmed_hypoxia > 0, na.rm = TRUE),
            sum(!mapped))
), "Literature coverage")

```

Coverage inside the region that actually matters - without the two NCBI fallbacks
the last two rows would be lost:

```{r pubmed_target_coverage, echo=FALSE}

cov_row <- function(label, sel) {
  s <- table(factor(master_table$pubmed_source[sel],
                    levels = levels(master_table$pubmed_source)))
  data.frame(Set = label, Genes = sum(sel), as.list(s), check.names = FALSE)
}

tab(rbind(
  cov_row("all HIF targets", !is.na(master_table$Target)),
  cov_row("Kelly-specific targets",
          !is.na(master_table$Target) & master_table$Hx_Baseline == "none" &
            !is.na(master_table$Hx_Baseline))
), "Entrez ID resolution within the target sets")

```

Sanity check - the genes with the most hypoxia literature are the ones you would
expect, which is what makes the column trustworthy:

```{r pubmed_sanity, echo=FALSE}

master_table %>%
  dplyr::arrange(dplyr::desc(pubmed_hypoxia)) %>%
  dplyr::select(symbol, pubmed_total, pubmed_hypoxia) %>%
  head(6) %>%
  tab("Best covered genes in the hypoxia literature")

```

## Candidates: strong, Kelly-specific, undescribed in hypoxia

HIF targets that respond only in Kelly, are strongly induced, and are established
genes (at least 20 publications) with **no** hypoxia literature at all.

```{r pubmed_candidates, echo=FALSE}

master_table %>%
  dplyr::filter(!is.na(Target), Hx_Baseline == "none", !is.na(symbol),
                pubmed_hypoxia == 0, pubmed_total >= 20) %>%
  dplyr::arrange(dplyr::desc(Kelly.Hx.vs.Nx.log2FoldChange)) %>%
  dplyr::mutate(log2FC = round(Kelly.Hx.vs.Nx.log2FoldChange, 2)) %>%
  dplyr::select(symbol, Target, pubmed_total, pubmed_hypoxia, log2FC) %>%
  head(12) %>%
  tab("Strongly induced Kelly-specific targets with no hypoxia literature")

```


# Clinical risk (HR)

`HR` is the hazard ratio from the high-risk neuroblastoma microarray cohort:
above 1 means high expression goes with worse outcome, below 1 with better.
The cutoff is the one already used in `4_manuscript/Figures.Rmd` - `HR > 1.5` or
`HR < 0.67` at `p_value_HR < 0.05`. The two bounds are reciprocals, so the
threshold is symmetric on the log scale, which is also the common convention for
dichotomising hazard ratios.

| Value | Meaning |
|---|---|
| `adverse` | HR > 1.5, p < 0.05 - high expression, worse prognosis |
| `protective` | HR < 0.67, p < 0.05 - high expression, better prognosis |
| `ns` | tested, but below the effect or significance threshold |
| `NA` | gene not present in the microarray cohort |

Two properties of this data are worth knowing before leaning on the column.
**The p-value does little of the filtering**: 58% of the tested genes fall below
0.05 and a BH correction only drops that from 14,759 to 13,699 genes, so the
selection is driven almost entirely by the HR magnitude. **The degenerate Cox
fits are harmless**: a handful of genes carry absurd ratios (TRPV5 at 1.1e29),
but those come with p close to 1 and are removed by the significance filter -
99.6% of all HR values sit inside [0.1, 10] anyway.

```{r hr_column}

hr_cut <- 1.5
hr_p   <- 0.05

master_table$HR_risk <- dplyr::case_when(
  is.na(master_table$HR) | is.na(master_table$p_value_HR) ~ NA_character_,
  master_table$p_value_HR < hr_p & master_table$HR > hr_cut     ~ "adverse",
  master_table$p_value_HR < hr_p & master_table$HR < 1 / hr_cut ~ "protective",
  TRUE                                                          ~ "ns"
)

master_table <- master_table %>% dplyr::relocate(HR_risk, .after = pubmed_source)

hr_f <- factor(ifelse(is.na(master_table$HR_risk), "no data", master_table$HR_risk),
               levels = c("adverse", "protective", "ns", "no data"))

tab(as.data.frame(table(HR_risk = hr_f)),
    sprintf("Clinical risk classification (HR > %.2f or < %.2f at p < %.2f)",
            hr_cut, 1 / hr_cut, hr_p))

```

## Risk within the HIF targets

```{r hr_targets}

tab(as.data.frame.matrix(table(Target = target_f, HR = hr_f)),
    "Target x HR_risk")

```

## Shortlist: Kelly-specific, adverse prognosis, undescribed in hypoxia

The three annotations combined - a target that responds only in Kelly, predicts
worse outcome in patients, and has no hypoxia literature yet.

```{r hr_shortlist, echo=FALSE}

master_table %>%
  dplyr::filter(!is.na(Target), Hx_Baseline == "none", HR_risk == "adverse",
                !is.na(symbol), pubmed_hypoxia == 0) %>%
  dplyr::arrange(dplyr::desc(Kelly.Hx.vs.Nx.log2FoldChange)) %>%
  dplyr::mutate(log2FC = round(Kelly.Hx.vs.Nx.log2FoldChange, 2),
                HR = round(HR, 2),
                p_HR = sprintf("%.1e", p_value_HR)) %>%
  dplyr::select(symbol, Target, log2FC, HR, p_HR, pubmed_total) %>%
  head(15) %>%
  tab("Kelly-specific, adverse prognosis, no hypoxia literature")

```


# Master counts table

The same table plus the per-sample normalized counts of the Kelly experiment.
Sample columns are relabelled `counts_genotype_condition_replicate`, so the
`counts_` prefix separates them from the statistics columns in one selection.

```{r build_master_counts_table}

sample_labels <- colData(dds_kelly) %>%
  as.data.frame() %>%
  tibble::rownames_to_column("sample_id") %>%
  dplyr::group_by(genotype, condition) %>%
  dplyr::mutate(replicate = dplyr::row_number()) %>%
  dplyr::ungroup() %>%
  dplyr::mutate(label = paste("counts", genotype, condition, replicate, sep = "_"))

norm_counts <- counts(dds_kelly, normalized = TRUE)
colnames(norm_counts) <- sample_labels$label[match(colnames(norm_counts), sample_labels$sample_id)]

counts_df <- norm_counts %>%
  as.data.frame() %>%
  tibble::rownames_to_column("Ensembl")

master_counts_table <- master_table %>% left_join(counts_df, by = "Ensembl")

count_cols <- grep("^counts_", colnames(master_counts_table), value = TRUE)

master_counts_table %>%
  dplyr::select(symbol, Target, Hx_Baseline, dplyr::all_of(head(count_cols, 4))) %>%
  head(5) %>%
  dplyr::mutate(dplyr::across(where(is.numeric), ~round(.x, 1))) %>%
  tab(sprintf("master_counts_table preview - %d columns total, %d of them counts",
              ncol(master_counts_table), length(count_cols)))

```


# Export

```{r export}

# Absolute paths via here(), so the exports always land in 6_XM_data_merge/ - knit
# sets the wd to this Rmd's folder, but running the chunks from the console uses
# the project root, which previously produced a second copy of both files there.
out_dir <- here::here("6_XM_data_merge")

openxlsx::write.xlsx(master_table,
                     file.path(out_dir, "master_table.xlsx"), rowNames = FALSE)
openxlsx::write.xlsx(master_counts_table,
                     file.path(out_dir, "master_counts_table.xlsx"), rowNames = FALSE)

```

```{r export_links, results='asis', echo=FALSE}

exports <- list(
  list(file = "master_table.xlsx",        obj = master_table,
       note = "statistics only"),
  list(file = "master_counts_table.xlsx", obj = master_counts_table,
       note = "statistics + per-sample normalized counts")
)

for (e in exports) {
  size_mb <- file.size(file.path(out_dir, e$file)) / 1024^2
  cat(sprintf("- [%s](%s) - %s genes x %d columns, %.1f MB (%s)\n",
              e$file, e$file, format(nrow(e$obj), big.mark = ","),
              ncol(e$obj), size_mb, e$note))
}

```
