---
title: "Master_Table_Dive"
output:
  github_document:
    html_preview: false
    toc: true
    toc_depth: 2
always_allow_html: true
editor_options:
  chunk_output_type: console
knit: (function(input_file, encoding) {
    rmarkdown::render(input_file,output_file= 'Dive.md')
        })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, echo = TRUE)

# Fold every code block into a collapsible <details> element. GitHub renders this
# natively, so the rendered document reads as a summary while the full code stays
# one click away.
knitr::knit_hooks$set(source = function(x, options) {
  paste0("\n<details>\n<summary>Code</summary>\n\n``` r\n",
         paste(x, collapse = "\n"), "\n```\n\n</details>\n")
})

# All summary output goes through kable, with the caption as a bold line *above*
# the table - kable's own caption ends up below it, where it reads as a stray
# sentence on GitHub.
tab <- function(x, caption = NULL) {
  md <- paste(knitr::kable(x), collapse = "\n")
  if (!is.null(caption)) md <- paste0("**", caption, "**\n\n", md)
  knitr::asis_output(paste0("\n", md, "\n"))
}
```

Follow-up analysis on the annotated master table built in
[Master_Table.Rmd](Master_Table.Rmd) (see [its rendered output](Readme.md) for how
the annotation columns are defined). This document starts by reducing the table
to the genes that are actually worth looking at. Code blocks are collapsed -
click *Code* to expand.

```{r load_packages}
library(dplyr)
library(tidyverse)
library(knitr)
library(ggplot2)
library(ggrepel)
library(openxlsx)

```


# Input

`master_counts_table.xlsx` carries everything: the DESeq2 contrasts, the
annotation columns, and the per-sample normalized counts. Reading 65 MB of Excel
on every knit is slow, so it is cached as an RDS and only re-read when the xlsx
is newer.

```{r load_master}

in_dir  <- here::here("6_XM_data_merge")
xlsx_in <- file.path(in_dir, "master_counts_table.xlsx")
rds_cache <- file.path(in_dir, "cache", "master_counts_table.rds")

dir.create(dirname(rds_cache), showWarnings = FALSE, recursive = TRUE)
if (file.exists(rds_cache) &&
    file.mtime(rds_cache) > file.mtime(xlsx_in)) {
  master <- readRDS(rds_cache)
} else {
  master <- as.data.frame(readxl::read_xlsx(xlsx_in))
  saveRDS(master, rds_cache)
}

count_cols <- grep("^counts_", colnames(master), value = TRUE)

tab(data.frame(
  Item = c("genes", "columns", "of them per-sample counts", "annotation columns"),
  n    = c(nrow(master), ncol(master), length(count_cols),
           sum(colnames(master) %in% c("Target", "Hx_Baseline", "pubmed_total",
                                       "pubmed_hypoxia", "pubmed_source", "HR_risk")))
), "master_counts_table as loaded")

```


# Reduction

Most of the table is not expressed: 61% of all genes have a `baseMean_Kelly` of
exactly zero. Expression is therefore the primary filter.

A gene is kept anyway if it has at least 50 publications. That rescue is there so
well-characterised genes stay visible as a reference even when they are silent in
this cell line - TLR2, HFE, NOD2, POSTN and DCN are all annotated HIF targets
that a pure expression cutoff would drop.

**The threshold is deliberately low**, because `pubmed_total` counts *curated
gene-to-publication links*, not mentions. NCBI creates those links mainly for
papers treating a gene as a molecular entity, which systematically undercounts
genes whose literature is old, physiological or biochemical:

| Gene | PubMed full-text search | `pubmed_total` |
|---|---:|---:|
| myoglobin / `MB` | 18,113 | 95 |
| hemoglobin / `HBB` | 286,325 | 876 |
| insulin / `INS` | 522,216 | 1,034 |
| `TP53` | - | 20,399 |

So the column measures how intensively a gene is studied *as a gene*, not how
famous it is. That is usually the more useful reading here - GAPDH scores 637
despite appearing in tens of thousands of papers as a loading control, and those
papers are not about GAPDH - but classic literature is underrepresented, so 50
links already marks a well-studied gene.

One consequence of Ensembl-to-Entrez being many-to-one: 1,691 symbols occupy more
than one row (NLRP2 has four Ensembl loci sharing Entrez 55655), and each row
carries the same literature count. Rows are genes as annotated here, so a count
of "genes above N publications" double-counts those. This comes from the
annotation itself, not from the fallback mapping - the duplication rate is 12.5%
for `annotation` and 13.2% for `symbol_map`.

Literature coverage is used **only** in that direction: it can keep a gene in, it
never throws one out. A strongly expressed but unstudied gene is a candidate, not
noise, and filtering on publication counts would preferentially discard exactly
those - one Kelly-specific target sits at `baseMean` 7,235 with no annotation at
all. `pubmed_total` and `pubmed_hypoxia` otherwise stay in the table as a ranking
aid.

```{r reduction}

min_basemean <- 10
min_pubmed   <- 50

expressed <- !is.na(master$baseMean_Kelly) & master$baseMean_Kelly >= min_basemean
rescued   <- !expressed & !is.na(master$pubmed_total) & master$pubmed_total >= min_pubmed
keep      <- expressed | rescued

dive <- master[keep, ]

# Why each gene is in, so the rescued ones can be excluded again downstream
# without recomputing the rule.
dive$dive_reason <- ifelse(expressed[keep], "expressed", "literature")
dive <- dive %>% dplyr::relocate(dive_reason, .after = symbol)

tab(data.frame(
  Step = c("master_counts_table",
           sprintf("baseMean_Kelly >= %d", min_basemean),
           sprintf("+ rescued: pubmed_total >= %d", min_pubmed)),
  Genes = c(nrow(master), sum(expressed), nrow(dive)),
  Share = sprintf("%.0f%%", 100 * c(nrow(master), sum(expressed), nrow(dive)) / nrow(master))
), "Reduction")

```

The rescue adds 11% to the table. Worth knowing before using it: 617 of the
rescued genes have a `baseMean` of exactly zero, so they enter as rows of pure
zero counts. That is informative as a reference - "this well-known gene is not
expressed in Kelly" - but they should be excluded from anything that consumes the
count columns, which `dive_reason` makes easy.

```{r rescue_detail, echo=FALSE}

tab(data.frame(
  baseMean = c("exactly 0", "0 - 1", "1 - 10"),
  Genes = c(sum(rescued & master$baseMean_Kelly == 0),
            sum(rescued & master$baseMean_Kelly > 0 & master$baseMean_Kelly < 1),
            sum(rescued & master$baseMean_Kelly >= 1 & master$baseMean_Kelly < min_basemean))
), sprintf("Expression of the %s literature-rescued genes",
           format(sum(rescued), big.mark = ",")))

```

What survives, by annotation. The reduction removes three quarters of the rows
but keeps the large majority of the annotated candidates, which is the point:

```{r reduction_effect, echo=FALSE}

keep_stats <- function(label, sel) {
  data.frame(Set = label,
             before      = sum(sel),
             expressed   = sum(sel & expressed),
             rescued     = sum(sel & rescued),
             after       = sum(sel & keep),
             kept        = sprintf("%.0f%%", 100 * sum(sel & keep) / max(sum(sel), 1)))
}

tab(rbind(
  keep_stats("all genes", rep(TRUE, nrow(master))),
  keep_stats("HIF targets", !is.na(master$Target)),
  keep_stats("Kelly-specific targets",
             !is.na(master$Target) & master$Hx_Baseline == "none" &
               !is.na(master$Hx_Baseline)),
  keep_stats("HR adverse", master$HR_risk == "adverse" & !is.na(master$HR_risk)),
  keep_stats("with hypoxia literature",
             !is.na(master$pubmed_hypoxia) & master$pubmed_hypoxia > 0)
), "Effect of the reduction per annotation set")

```

The best-studied HIF targets that a pure expression cutoff would have dropped -
all of them now retained by the rescue:

```{r rescued_examples, echo=FALSE}

master[rescued & !is.na(master$Target), ] %>%
  dplyr::arrange(dplyr::desc(pubmed_total)) %>%
  dplyr::mutate(baseMean = round(baseMean_Kelly, 2)) %>%
  dplyr::select(symbol, Target, baseMean, pubmed_total, pubmed_hypoxia) %>%
  head(8) %>%
  tab("HIF targets kept by the literature rescue")

```

Still dropped: annotated targets that are both silent and barely studied. These
are the ones the filter is meant to catch.

```{r dropped_examples, echo=FALSE}

master[!keep & !is.na(master$Target), ] %>%
  dplyr::arrange(dplyr::desc(pubmed_total)) %>%
  dplyr::mutate(baseMean = round(baseMean_Kelly, 2)) %>%
  dplyr::select(symbol, Target, baseMean, pubmed_total, pubmed_hypoxia) %>%
  head(6) %>%
  tab("Best-studied HIF targets still removed")

```


# Reduced table

```{r dive_preview, echo=FALSE}

dive %>%
  dplyr::filter(!is.na(Target), Hx_Baseline == "none", !is.na(symbol)) %>%
  dplyr::arrange(dplyr::desc(Kelly.Hx.vs.Nx.log2FoldChange)) %>%
  dplyr::mutate(log2FC = round(Kelly.Hx.vs.Nx.log2FoldChange, 2),
                baseMean = round(baseMean_Kelly, 0)) %>%
  dplyr::select(symbol, Target, baseMean, log2FC, HR_risk,
                pubmed_total, pubmed_hypoxia) %>%
  head(10) %>%
  tab("Top Kelly-specific targets in the reduced table")

```

# Risk versus HIF dependence

One panel per HIF, showing every target of that HIF against two axes:

- **x** - clinical risk as `log2(HR)`. Left of zero the gene is protective, right
  of it adverse. The dotted lines mark the 1.5 / 0.67 thresholds used for
  `HR_risk`.
- **y** - HIF dependence, the interaction term negated so that *up* means *more
  dependent on this HIF*. Capped at 8; nine genes sit above it and are drawn at
  the cap.

Four further variables ride along:

| Channel | Meaning |
|---|---|
| Hue | red / teal = Kelly-specific, grey = also hypoxic in the baseline lines, pale grey = no baseline data |
| Saturation | expression (`baseMean_Kelly`) |
| Size | publications (`pubmed_total`) |
| Amber ring | at least 50 publications but **none** on hypoxia |

The ring rule deserves a note. Simply marking "few hypoxia papers" would highlight
60% of all targets and say nothing - most genes have no hypoxia literature because
most genes are not well studied at all. Requiring 50 publications first makes the
absence informative: a gene people have worked on, that nobody has looked at under
hypoxia. That combination - a **large, ringed, coloured** point in the upper right
- is the shortlist in visual form.

```{r risk_vs_dependence, fig.width=11, fig.height=6.4, dpi=150}

COL <- c("Kelly-specific (HIF1A)" = "#C0392B", "Kelly-specific (HIF2A)" = "#16A085",
         "also Hx in SK-N-AS / SH-SY5Y" = "#B4B4B4", "no baseline data" = "#E4E4E4")
COL_RING <- "#E8B006"
ring_min_total <- 50
y_cap <- 8

prep <- function(grp, ycol, panel, tgt_lab) {
  dive %>%
    dplyr::filter(Target %in% grp, !is.na(HR), HR >= 0.1, HR <= 10) %>%
    dplyr::transmute(
      panel = panel, symbol,
      log2HR = log2(HR), dep = pmin(-.data[[ycol]], y_cap),
      cls = dplyr::case_when(is.na(Hx_Baseline)     ~ "no baseline data",
                             Hx_Baseline == "none"  ~ tgt_lab,
                             TRUE ~ "also Hx in SK-N-AS / SH-SY5Y"),
      expr = log10(baseMean_Kelly + 1),
      pm   = log10(pubmed_total + 1),
      ring = !is.na(pubmed_hypoxia) & pubmed_hypoxia == 0 &
             !is.na(pubmed_total) & pubmed_total >= ring_min_total,
      kelly = !is.na(Hx_Baseline) & Hx_Baseline == "none")
}

pdat <- dplyr::bind_rows(
  prep(c("Hif1a", "Hif1a_Hif2a"), "Hif1aHxNx.vs.KellyHxNx.log2FoldChange",
       "HIF1A targets", "Kelly-specific (HIF1A)"),
  prep(c("Hif2a", "Hif1a_Hif2a"), "Hif2aHxNx.vs.KellyHxNx.log2FoldChange",
       "HIF2A targets", "Kelly-specific (HIF2A)"))
pdat$cls   <- factor(pdat$cls, levels = names(COL))
pdat$panel <- factor(pdat$panel, levels = c("HIF1A targets", "HIF2A targets"))

lab <- pdat %>% dplyr::filter(kelly, ring, log2HR > log2(1.5)) %>%
  dplyr::group_by(panel) %>% dplyr::arrange(dplyr::desc(dep)) %>%
  dplyr::slice_head(n = 7) %>% dplyr::ungroup()

ggplot(pdat, aes(log2HR, dep)) +
  geom_vline(xintercept = c(log2(1 / 1.5), log2(1.5)), colour = "grey85",
             linetype = "dotted", linewidth = 0.35) +
  geom_point(aes(colour = cls, size = pm, alpha = expr)) +
  geom_point(data = dplyr::filter(pdat, ring),
             aes(size = pm, shape = "no hypoxia literature\n(>=50 papers, 0 on hypoxia)"),
             fill = NA, colour = COL_RING, stroke = 0.65) +
  ggrepel::geom_text_repel(data = lab, aes(label = symbol), size = 2.5,
                           colour = "grey15", box.padding = 0.45, max.overlaps = 40,
                           segment.size = 0.2, segment.colour = "grey55",
                           min.segment.length = 0) +
  facet_wrap(~panel) +
  scale_colour_manual(values = COL, name = NULL, guide = guide_legend(
    order = 1, ncol = 2, override.aes = list(size = 3.2, alpha = 1))) +
  scale_size_continuous(range = c(1, 4.6), name = "publications",
    breaks = log10(c(10, 100, 1000) + 1), labels = c("10", "100", "1000"),
    guide = guide_legend(order = 2, override.aes = list(colour = "grey45", alpha = 1))) +
  scale_alpha_continuous(range = c(0.35, 1), name = "expression (baseMean)",
    breaks = log10(c(10, 1000, 100000) + 1), labels = c("10", "1k", "100k"),
    guide = guide_legend(order = 3, override.aes = list(size = 3.2, colour = "grey25"))) +
  scale_shape_manual(values = 21, name = NULL, guide = guide_legend(
    order = 4, override.aes = list(size = 3.4, colour = COL_RING, stroke = 0.9))) +
  labs(x = "clinical risk    log2(hazard ratio)    ← protective | adverse →",
       y = sprintf("HIF dependence    −log2FC interaction (capped at %d)", y_cap)) +
  theme_minimal(base_size = 9) +
  theme(panel.grid.minor = element_blank(),
        panel.grid.major = element_line(linewidth = 0.2, colour = "grey94"),
        strip.text = element_text(face = "bold", size = 10),
        legend.position = "bottom", legend.box = "horizontal",
        legend.key.height = unit(0.75, "lines"),
        legend.title = element_text(size = 8), legend.text = element_text(size = 7.5))

```

```{r plot_counts, echo=FALSE}

n_tot <- function(grp) sum(dive$Target %in% grp)
tab(data.frame(
  Panel = c("HIF1A targets", "HIF2A targets"),
  in_dive = c(n_tot(c("Hif1a", "Hif1a_Hif2a")), n_tot(c("Hif2a", "Hif1a_Hif2a"))),
  plotted = as.integer(table(pdat$panel)),
  Kelly_specific = as.integer(tapply(pdat$kelly, pdat$panel, sum)),
  ringed = as.integer(tapply(pdat$ring, pdat$panel, sum))
), "Genes per panel - the difference to `in_dive` are genes without a usable hazard ratio")

```


```{r export}

out_file <- here::here("6_XM_data_merge", "master_table_dive.xlsx")
openxlsx::write.xlsx(dive, out_file, rowNames = FALSE)

```

```{r export_link, results='asis', echo=FALSE}

cat(sprintf("- [master_table_dive.xlsx](master_table_dive.xlsx) - %s genes x %d columns, %.1f MB\n",
            format(nrow(dive), big.mark = ","), ncol(dive),
            file.size(out_file) / 1024^2))

```
